Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-07-28 paper/main.tex
related work tp and sw

\IEEEPARstart{V}isualizations are expected to be a faithful translation of data, meaning that inherent structure in the data should be present in the visualization; therefore the components of visualization software libraries that translate data to graphics are also expected to be structure preserving. This inherent structure can be quite complex, such as when the data is high dimensional, multivariate, deeply nested, or encoded in a distributed manner. We propose that we can rigorously specify what structure the library components are expected to preserve using category theory. We also propose that these specifications can generalize to most structure types by extending previous work on encoding  inherent structure in a consistent manner using mathematical structures from algebraic topology. These algebraic structures allow us to encode inherent structure as programmatic types, and the use of category theory facilitates developing a functional design framework that describes composability in terms of functional algebraic operations. The typing system and composability inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects by design\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library, but can also be integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework provides also provides guidence for the construction and testing of structure preserving visualization library components.

@@

There is an assumption in visualization that the connectivity of the elements in a dataset are preserved in the visual representation; roughly topology provides a method for encoding this connectivity \cite{wilkinsonGrammarGraphics2005}. Bertin gives the examples of discrete data values should be represented by discrete entities such as points, quantitative continuous data by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011}. In \autoref{fig:related-work:continuity:ktypes} the records of the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} weather station table can be considered disconnected from each other and therefore have a 0D continuity. This means each distinct row can be reduced to a discrete point and it is expected  that each rows will be encoded as a discrete visual element, such as a standalone scatter point. Each station can also be considered a time series of weather observation since temperature and pressure are sampled from a 1D continuous measurement space. The connectivity of these observations can be modeled as an interval and the data is encoded as a line since that visual element is also 1D continuous. The measurements at set of stations on any given day is a sparse sampling of a 2D geospatial grid and the connectivity can be modeled as a 2D plane. Instead of interpolation, which is how continuity was preserved in the timeseries plot, in this instance the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved. Encoding connectivity using the formalism of topological spaces allows us to encode the connectivity of the data in a uniform manner such that we can then generalize the preservation of that connectivity in the visual representation rather than constructing a condition specifically for each type of topology.

Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi and by Troy and M\"{o}ller \cite{chiTaxonomyVisualizationTechniques2000,toryRethinkingVisualizationHighlevel2004}. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner. Since topological continuity is generalizable, we propose the construction of topology types such that data can explicitly state its topology type and visual algorithms can check that this type matches their assumptions. The typing system proposed in this paper builds on Butler's proposal of using a mathematical structure called fiber bundles as an abstract data representation in visualization \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}. We sketch out fiber bundles in \autoref{sec:atct:fiber-bundles}, but his work provides a thorough introduction to bundles for visualization practitioners.

@@

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are, where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomatingDesignGraphical1986}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. Libraries that provide small buildable components \cite{wongsuphasawatNavigatingWideWorld2021}, such as components for making boxes or translating data values to colors, mostly embed their connectivity assumptions in their visual algorithms and allow the user choose almost any component, regardless of structure preservation. Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the space of choices to algorithms that are appropriate for the data. 
Buildable component libraries such as matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} support a wide variety of data types, but at the cost of visual algorithms with  parameters that are tuned to the expected data in a way that leaves the over all library feeling uncohesive. VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, but does so in a non-uniform manner. On the other hand, the topology of a table as a set of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanaties data ImagePlot\cite{studiesCulturevisImageplot2021} plugin assumed that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph like structure. These domain specific libraries tend to benefit from more cohesive apis, but at the cost of a much more limited set of visualization algorithms than the building block libraries.

>>>>

\IEEEPARstart{V}isualizations are expected to be a faithful translation of data, meaning that inherent structure in the data should be present in the visualization; therefore the components of visualization software libraries that translate data to graphics are also expected to be structure preserving. This inherent structure can be quite complex, such as when the data is high dimensional, multivariate, deeply nested, or encoded in a distributed manner. We propose that we can rigorously specify what structure the library components are expected to preserve using category theory. We also propose that these specifications can generalize to most structure types by extending previous work on encoding  inherent structure in a consistent manner using mathematical structures from algebraic topology. These algebraic structures allow us to encode inherent structure as programmatic types, and the use of category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects by design\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library, but can also be integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework provides also provides guidance for the construction and testing of structure preserving visualization library components.

@@

There is an assumption in visualization that the connectivity of the elements in a dataset are preserved in the visual representation; roughly topology provides a method for encoding this connectivity \cite{wilkinsonGrammarGraphics2005}. Bertin gives the examples of discrete data values should be represented by discrete entities such as points, quantitative continuous data by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011}. In \autoref{fig:related-work:continuity:ktypes} the records of the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} weather station table can be considered disconnected from each other and therefore have a 0D continuity. This means each distinct row can be reduced to a discrete point and each row will be encoded as a discrete visual element, such as a standalone scatter point. Each station can also be considered a time series of weather observation since temperature and pressure are sampled from a 1D continuous measurement space. The connectivity of these observations can be modeled as an interval and the data is encoded as a line since that visual element is also 1D continuous. The measurements at set of stations on any given day is a sparse sampling of a 2D geospatial grid and the connectivity can be modeled as a 2D plane. Instead of interpolation, which is how continuity was preserved in the timeseries plot, in this instance the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved. Encoding connectivity using the formalism of topological spaces allows us to encode the connectivity of the data in a uniform manner such that we can then generalize the preservation of that connectivity in the visual representation rather than constructing a condition specifically for each type of topological structure.

Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi\cite{chiTaxonomyVisualizationTechniques2000} and by Troy and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004}. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner. Since topological continuity is generalizable, we propose the construction of topology types such that data can explicitly state its topology type and visual algorithms can check that this type matches their assumptions. The typing system proposed in this paper builds on Butler's proposal of using a mathematical structure called fiber bundles as an abstract data representation in visualization \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}. We extend Butler's work because he discusses how bundles can be used to encode data in a consistent manner for a number of different \note{arbitrary?} topological and field structures. 
 We sketch out fiber bundles in \autoref{sec:atct:fiber-bundles}, but his work provides a thorough introduction to bundles for visualization practitioners.

@@

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are, where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomatingDesignGraphical1986}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. For example, connectivity assumptions tend to be embedded in each of the visual algorithms of `building block` libraries, a term coined by by Wongsuphasawat \cite{wongsuphasawatNavigatingWideWorld2021} to describe libraries that provide modular components for building elements of a visualization, such as functions for making boxes or translating data values to colors. In building block libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} the connectivity assumptions lead to very different interfaces; for example in Matplotlib methods for updating the data in a plot differ across the various plotting objects and the parameter list is inconsistent as to which parameters can be applied per data point or for the whole element, and in D3 the API for network visualizers is mostly distinct from the rest of the modules. While VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, and therefore can embed that information in its visual algorithms, it does so in a non-uniform manner. On the other hand, domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the interface to choices that that are appropriate for the domain. For example, a tabluar topological structure of discret rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanities ImagePlot\cite{studiesCulturevisImageplot2021} plugin assume that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph-like structure. By assuming the structure of their data, these domain specific libraries can provide more cohesive interfaces for a much more limited set of visualization algorithms than the building block libraries offer.
